Endogenous Macrodynamics in Algorithmic Recourse

### 2022-08-09 dev/paper/paper.tex
some work on rephrasing narrative in intro

\begin{document}
\maketitle
\begin{abstract}
Existing work on Counterfactual Explanations (CE) and Algorithmic
Recourse (AR) has largely been limited to the static setting: given some
classifier we are interested in finding close, actionable, realistic,
sparse, diverse and ideally causally founded counterfactuals. The
ability of CE to handle dynamics like data and model drift remains a
largely unexplored research challenge at this point. Only one recent
work considers the implications of exogenous domain and model shifts.
This project instead focuses on endogenous dynamics, that is shifts that
occur when AR is actually implemented by a proportion of individuals.
Early findings suggest that the involved shifts may be large with
important implications on the validity of AR and the overall
characteristics of the sample population.
\end{abstract}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced, boxrule=0pt, frame hidden, sharp corners, breakable]}{\end{tcolorbox}}\fi

@@

Figure~\ref{fig-poc} illustrates this idea for a binary problem
involving a probabilistic classifier and the counterfactual generator
proposed by \protect\hyperlink{ref-wachter2017counterfactual}{{[}5{]}}:
the implementation of AR for a subset of individuals leads to a domain
shift (b), which in turn triggers a model shift (c). As this game of
implementing AR and updating the classifier is repeated, the decision
boundary moves away from training samples that were originally in the
target class (d). We refer to these types of dynamics as
\textbf{endogenous} because they are induced by the implementation of
recourse itself.

@@

We think that these types of endogenous dynamics may be problematic.
Firstly, model shifts may inadvertently change classification outcomes
for individuals who never received and implemented recourse. Secondly
and relatedly, we observe in Figure~\ref{fig-poc} that as the decision
boundary moves in the direction of the non-target class, counterfactual
paths become shorter: in the consumer credit example, individuals that
previously would have been denied credit based on their input features
are suddenly considered as creditworthy. Average default risk across all
borrowers can therefore be expected to increase. Conversely, lenders
that anticipate such dynamics may choose to deny credit to individuals
that have implemented AR, thereby compromising the validity of AR.

To the best of our knowledge this is the first work investigating
endogenous dynamics in AR. Our contributions to the state of knowledge
are two-fold. Firstly, we introduce an experimental framework extending
previous work by \protect\hyperlink{ref-pawelczyk2021carla}{{[}4{]}},
which allows for benchmarking different counterfactual generators in
terms of the endogenous domain and model shifts they introduce. To this
end we propose a number of novel evaluation metrics. Secondly, we use
this framework to provide the first in-depth analysis of endogenous
recourse dynamics induced by various popular counterfactual generators
including \protect\hyperlink{ref-wachter2017counterfactual}{{[}5{]}},
\protect\hyperlink{ref-joshi2019towards}{{[}6{]}},
\protect\hyperlink{ref-mothilal2020explaining}{{[}7{]}} and
\protect\hyperlink{ref-antoran2020getting}{{[}8{]}}.

We find that \ldots{}

The paper is structured as follows: Section~\ref{sec-related} places our
work in the broader context of related literature.
Section~\ref{sec-method} presents our methodology and data.
Section~\ref{sec-empirical} presents our empirical findings which are
then discussed in Section~\ref{sec-discussion}. We also point to some of
the limitations or our work as well as avenues for future research in
Section~\ref{sec-limit}. Finally, Section~\ref{sec-conclusion}
concludes.

@@

Many approaches for the generation of algorithmic recourse have been
described in the literature. An October 2020 survey by Karimi et
al.~layed out 60 algorithms that have been proposed since 2014
\protect\hyperlink{ref-karimi2020survey}{{[}9{]}}. Another survey
published around the same time by Verma et al.~described 29 algorithms
\protect\hyperlink{ref-verma2020counterfactual}{{[}10{]}}. Different
approaches vary primarily in terms of the objective functions they
impose, how they optimize said objective (from brute force through
gradient-based approaches to graph traversal algorithms), and how the
ensure that certain requirements for CE are met. Regarding the latter,
the literature has produced an extensive list of desiderata each
addressing different needs. To name but a few, we are interested in
generating counterfactuals that close
\protect\hyperlink{ref-wachter2017counterfactual}{{[}5{]}}, actionable
\protect\hyperlink{ref-ustun2019actionable}{{[}11{]}}, realistic
\protect\hyperlink{ref-schut2021generating}{{[}12{]}}, sparse, diverse
\protect\hyperlink{ref-mothilal2020explaining}{{[}7{]}} and if possible
causally founded
\protect\hyperlink{ref-karimi2021algorithmic}{{[}13{]}}.

Efforts so far have largely been directed at improving the quality of
counterfactual explanations within a static context: given some
pre-trained classifier \(M: \mathcal{X} \mapsto \mathcal{Y}\) we are
interested in generating one or multiple meaningful counterfactual
explanations for some individual characterized by \(x_i\). The ability
of counterfactual explanations to handle dynamics like data and model
shifts remains a largely unexplored research challenge at this point
\protect\hyperlink{ref-verma2020counterfactual}{{[}10{]}}. We have been
able to identify only one recent work that considers the implications of
\textbf{exogenous} domain and model shifts in the context of AR
\protect\hyperlink{ref-upadhyay2021towards}{{[}14{]}}. Exogenous shifts
are strictly of external origin. For example, they might stem from data
correction, temporal shifts or geospatial changes
\protect\hyperlink{ref-upadhyay2021towards}{{[}14{]}}. The authors of
\protect\hyperlink{ref-upadhyay2021towards}{{[}14{]}} propose framework
for algorithmic recourse (ROAR) that evidently improves robustness to
such exogenous shifts.

@@

Much attention has been paid to the detection of dataset shifts --
situations where the distribution of data changes over time. Rabanser et
al.~suggest a framework to detect data drift from a minimal number of
samples through the application of two-sample tests
\protect\hyperlink{ref-rabanser2019failing}{{[}15{]}}. This task is a
generalization of the anomaly detection problem for large datasets,
which aims to answer the question if two sets of samples could have been
generated from the same probability distribution. Numerous approaches to
anomaly detection have been summarized
\protect\hyperlink{ref-chandola2009anomaly}{{[}16{]}}. Another
well-established research topic is that of concept drift -- situations
where external variables influence the patterns between the input and
the output of a model
\protect\hyperlink{ref-widmer1996learning}{{[}17{]}}. For instance, Gama
et al.~offer a review of the adaptive learning techniques which can
handle concept drift \protect\hyperlink{ref-gama2014survey}{{[}18{]}}.
Less previous work is available on the related topic of model drift -
changes in model performance over time. Nelson et al.~review how
resistant different machine learning models are to the model drift
\protect\hyperlink{ref-nelson2015evaluating}{{[}19{]}}. Ackerman et
al.~offer a method to detect changes in model performance when ground
truth is not available
\protect\hyperlink{ref-ackerman2021machine}{{[}20{]}}.

In the context of algorithmic recourse, domain and model shifts were
first brought up by the authors behind ROAR
\protect\hyperlink{ref-upadhyay2021towards}{{[}14{]}}. In their work
they refer to model shifts as simply any perturbation \(\Delta\) to the
parameters of the model in question: \(M\). While this also sets the
baseline for our analysis here, it is worth noting that in
\protect\hyperlink{ref-upadhyay2021towards}{{[}14{]}} these
perturbations are mechanically introduced. In contrast we are interested
in quantifying model shifts that arise endogenously as part of a dynamic
recourse process. In addition to quantifying the magnitude of shifts
\(\Delta\), we aim to also analyse the characteristics of changes to the
model, such as the position of the decision boundary and the overall
decisiveness of the model. We have not been able to identify previous
work on this topic.

@@

Despite the large and growing number of approaches to counterfactual
search there have been surprisingly few benchmark studies that compare
different methodologies. This may be partially due to limited software
availability in this space. Recent work has started to address this gap:
firstly, \protect\hyperlink{ref-de2021framework}{{[}21{]}} run a large
benchmarking study using different algorithmic aproaches and numerous
tabular datasets; secondly,
\protect\hyperlink{ref-pawelczyk2021carla}{{[}4{]}} introduce a Python
framework that can be used to apply and benchmark different methodology;
finally,
\protect\hyperlink{ref-altmeyer2022CounterfactualExplanations}{{[}22{]}}
provides an extensible and language-agnostic implementation in Julia.
For the experiments conducted in this work we relied on the latter.

@@

Here \(\mathbf{s}^\prime=\left\{s_k^\prime\right\}_K\) is the stacked
\(K\)-dimensional array of counterfactual states and
\(f: \mathcal{S} \mapsto \mathcal{X}\) maps from the counterfactual
state space to the feature space. In the case of the baseline
counterfactual generator
\protect\hyperlink{ref-wachter2017counterfactual}{{[}5{]}} \(f\) is just
the idendity function and the number of counterfactuals \(K\) is equal
to one. This generator, which we shall refer to as \textbf{Wachter} in
the following, shall serve as the baseline against which all other
gradient-based methodologies will be compared. In particular, we include
include the following generator in our benchmarking exercises: REVISE
\protect\hyperlink{ref-joshi2019towards}{{[}6{]}}, CLUE
\protect\hyperlink{ref-antoran2020getting}{{[}8{]}}, DICE
\protect\hyperlink{ref-mothilal2020explaining}{{[}7{]}} and a greedy
approach that relies on probabilistic models
\protect\hyperlink{ref-schut2021generating}{{[}12{]}}.

Both \textbf{REVISE} and \textbf{CLUE} search counterfactuals in some
latent embedding \(S \subset \mathcal{S}\) instead of the feature space
directly. The latent embedding is learned by a separate generative model
that is tasked with learning the data generating process (DGP) of \(X\).
In this case \(f\) in Equation~\ref{eq-general} corresponds to the
decoder part of the generative model, in other words the deterministic
function that maps back from the latent embedding to the feature space.
Provided the generative model is well-specified, traversing the latent
embedding typically results in realistic and plausible counterfactuals,
because they are implicitly generated by the (learned) DGP
\protect\hyperlink{ref-joshi2019towards}{{[}6{]}}. CLUE distinguishes
itself from REVISE and other counterfactual generators in that it aims
to minimize the predictive uncertainty of the model in question \(M\).
To quantify predictive uncertainty the authors rely on entropy estimates
for probabilistic models. The \textbf{Greedy} approach proposed by
\protect\hyperlink{ref-schut2021generating}{{[}12{]}} also works with
the subclass of models \(\tilde{\mathcal{M}}\subset\mathcal{M}\) that
can produce predictive uncertainty estimates. The authors show that in
this setting the complexity penalty \(h(\cdot)\) in
Equation~\ref{eq-general} is redundant and meaningful counterfactuals
can be generated in a fast and efficient manner through a modified
Jacobian-based Saliency Map Attack (JSMA). Finally, \textbf{DICE}
distinguishes itself from all other generators considered here in that
it aims to generate a diverse set of \(K>1\) counterfactuals. To this
end the authors use a complexity penalty \(h(\mathbf{s}^\prime)\) that
favours diverse outcomes, in the sense that \(s_1, ... , s_K\) look as
different from each other as possible.

@@

We have chosen to work with both synthetic and real-world datasets.
Using synthetic data allows us to impose distributional properties that
may affect the resulting recourse dynamics. Following
\protect\hyperlink{ref-upadhyay2021towards}{{[}14{]}} we generate
synthetic data in \(\mathbb{R}^2\) to also allow for a visual
interpretation of the results. Real-world data is used in order to
assess if endogenous dynamics also occur in higher-dimensional settings.

@@

Ex-ante we expect to see that Wachter will create a new cluster of
counterfactual instances in the proximity of the initial decision
boundary. Thus, the choice of a black-box model may have an impact on
the paths of the recourse. For generators that use latent space search
(\protect\hyperlink{ref-joshi2019towards}{{[}6{]}},
\protect\hyperlink{ref-antoran2020getting}{{[}8{]}}) or rely on (and
have access to) probabilistic models
(\protect\hyperlink{ref-antoran2020getting}{{[}8{]}},
\protect\hyperlink{ref-schut2021generating}{{[}12{]}}) we expect that
counterfactuals will end up in regions of the target domain that are
densely populated by training samples. Finally, we expect to see the
counterfactuals generated by DiCE to be uniformly spread around the
feature space inside the target class.

@@

As our baseline for quantifying model shifts we measure perturbations to
the model parameters at each point in time \(t\) following
\protect\hyperlink{ref-upadhyay2021towards}{{[}14{]}}. We define
\(\Delta=||\theta_{t+1}-\theta_{t}||^2\), that is the euclidean distance
between the vectors of parameters before and after retraining the model
\(M\). We shall refer to this baseline metric simply as
\textbf{Perturbations}.

@@

The experimental setup proposed here is designed to mimic a real-world
recourse process in a simple fashion. In practice, models are in fact
updated on a regular basis
\protect\hyperlink{ref-upadhyay2021towards}{{[}14{]}}. We also find it
plausible to assume that the implementation of recourse happens
periodically for different individuals, rather that all at once at time
\(t=0\). That being said, our experimental design is a vast
over-simplification of potential real-world scenarios. In practice, any
endogenous shifts that may occur can be expected to be entangled with
exogenous shifts of the nature investigated in
\protect\hyperlink{ref-upadhyay2021towards}{{[}14{]}}. We also make
implicit assumptions about the utility functions of the involved agents
that may well be too simple: individuals seeking recourse are assumed to
always implement the proposed counterfactual explanations; conversely,
the agent in charge of the model \(M\) is assumed to always treat
individuals that have implemented valid recourse as if they were truly
now in the target class. Relating this back to the consumer credit
example, we assume that the would-be borrowers are always willing and
able to implement recourse and the bank is always willing to provide
credit as would-be borrowers move across the decision boundary. In
practice it is doubtful that agents behave according to such simple
rules. Nonetheless, we think that our simple framework offers a starting
point for future work on recourse dynamics (both endogenous and
exogenous dynamics).

@@

\leavevmode\vadjust pre{\hypertarget{ref-joshi2019towards}{}}%
\CSLLeftMargin{{[}6{]} }%
\CSLRightInline{S. Joshi, O. Koyejo, W. Vijitbenjaronk, B. Kim, and J.
Ghosh, {``Towards realistic individual recourse and actionable
explanations in black-box decision making systems,''} \emph{arXiv
preprint arXiv:1907.09615}, 2019.}

\leavevmode\vadjust pre{\hypertarget{ref-mothilal2020explaining}{}}%
\CSLLeftMargin{{[}7{]} }%
\CSLRightInline{R. K. Mothilal, A. Sharma, and C. Tan, {``Explaining
machine learning classifiers through diverse counterfactual
explanations,''} in \emph{Proceedings of the 2020 conference on
fairness, accountability, and transparency}, 2020, pp. 607--617.}

\leavevmode\vadjust pre{\hypertarget{ref-antoran2020getting}{}}%
\CSLLeftMargin{{[}8{]} }%
\CSLRightInline{J. Antorán, U. Bhatt, T. Adel, A. Weller, and J. M.
Hernández-Lobato, {``Getting a clue: A method for explaining uncertainty
estimates,''} \emph{arXiv preprint arXiv:2006.06848}, 2020.}

\leavevmode\vadjust pre{\hypertarget{ref-karimi2020survey}{}}%
\CSLLeftMargin{{[}9{]} }%
\CSLRightInline{A.-H. Karimi, G. Barthe, B. Schölkopf, and I. Valera,
{``A survey of algorithmic recourse: Definitions, formulations,
solutions, and prospects,''} \emph{arXiv preprint arXiv:2010.04050},
2020.}

\leavevmode\vadjust pre{\hypertarget{ref-verma2020counterfactual}{}}%
\CSLLeftMargin{{[}10{]} }%
\CSLRightInline{S. Verma, J. Dickerson, and K. Hines, {``Counterfactual
explanations for machine learning: A review,''} \emph{arXiv preprint
arXiv:2010.10596}, 2020.}

\leavevmode\vadjust pre{\hypertarget{ref-ustun2019actionable}{}}%
\CSLLeftMargin{{[}11{]} }%
\CSLRightInline{B. Ustun, A. Spangher, and Y. Liu, {``Actionable
recourse in linear classification,''} in \emph{Proceedings of the
conference on fairness, accountability, and transparency}, 2019, pp.
10--19.}

\leavevmode\vadjust pre{\hypertarget{ref-schut2021generating}{}}%
\CSLLeftMargin{{[}12{]} }%
\CSLRightInline{L. Schut \emph{et al.}, {``Generating interpretable
counterfactual explanations by implicit minimisation of epistemic and
aleatoric uncertainties,''} in \emph{International conference on
artificial intelligence and statistics}, 2021, pp. 1756--1764.}

\leavevmode\vadjust pre{\hypertarget{ref-karimi2021algorithmic}{}}%
\CSLLeftMargin{{[}13{]} }%
\CSLRightInline{A.-H. Karimi, B. Schölkopf, and I. Valera,
{``Algorithmic recourse: From counterfactual explanations to
interventions,''} in \emph{Proceedings of the 2021 ACM conference on
fairness, accountability, and transparency}, 2021, pp. 353--362.}

\leavevmode\vadjust pre{\hypertarget{ref-upadhyay2021towards}{}}%
\CSLLeftMargin{{[}14{]} }%
\CSLRightInline{S. Upadhyay, S. Joshi, and H. Lakkaraju, {``Towards
robust and reliable algorithmic recourse,''} \emph{arXiv preprint
arXiv:2102.13620}, 2021.}

>>>>

\begin{document}
\maketitle
\begin{abstract}
Existing work on Counterfactual Explanations (CE) and Algorithmic
Recourse (AR) has largely been limited to the static setting: given some
classifier we are interested in finding close, actionable, realistic,
sparse, diverse and ideally causally founded counterfactuals. The
ability of CE to handle dynamics like data and model drift remains a
largely unexplored research challenge at this point. Only one recent
work considers the implications of exogenous domain and model shifts.
This project instead focuses on endogenous dynamics, that is shifts that
occur when AR is actually implemented by a proportion of individuals.
Early findings suggest that the involved shifts may be large with
important implications on the validity of AR and the overall
characteristics of the sample population.
\end{abstract}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, sharp corners, enhanced, borderline west={3pt}{0pt}{shadecolor}, breakable, boxrule=0pt, frame hidden]}{\end{tcolorbox}}\fi

@@

Research on algorithmic recourse has also so far typically addressed the
issue from the perspective of one single individual and has indeed been
referred to as \textbf{individual recourse} in some places. Arguably
though, most real-world applications that warrant algorithmic recourse
involve potentially large groups of individuals typically competing for
scarce resources. Our work demonstrates that in such scenarios, choices
made by or for one single individual are likely to affect the broader
collective of individuals in ways that current approaches to AR fail to
account for. More specifically, we argue that a strict focus on
minimizing the private costs faced by individuals may be too narrow an
objective.

Figure~\ref{fig-poc} illustrates this idea for a binary problem
involving a probabilistic classifier and the counterfactual generator
proposed by \protect\hyperlink{ref-wachter2017counterfactual}{{[}5{]}}:
the implementation of AR for a subset of individuals leads to a domain
shift (b), which in turn triggers a model shift (c). As this game of
implementing AR and updating the classifier is repeated, the decision
boundary moves away from training samples that were originally in the
target class (d). We refer to these types of dynamics as
\textbf{endogenous} because they are induced by the implementation of
recourse itself. The term \textbf{macrodynamics} is borrowed from the
economics literature and used to describe processes involving whole
groups or societies.

@@

We think that these types of endogenous dynamics may be problematic and
warrant our attention. Firstly, model shifts may inadvertently change
classification outcomes for individuals who never received and
implemented recourse. Secondly and relatedly, we observe in
Figure~\ref{fig-poc} that as the decision boundary moves in the
direction of the non-target class, counterfactual paths become shorter:
in the consumer credit example, individuals that previously would have
been denied credit based on their input features are suddenly considered
as creditworthy. Average default risk across all borrowers can therefore
be expected to increase. Conversely, lenders that anticipate such
dynamics may choose to refrain from offering recourse (and hence credit)
to more than just a tiny share of individuals. In that latter and
perhaps more likely scenario, the probability of being offered recourse
decreases with every individual that implements recourse: in other
words, the actions of first-movers exert a negative externality on
future would-be borrowers.

To the best of our knowledge this is the first work investigating
endogenous macrodynamics in AR. Our contributions to the state of
knowledge are as follows: firstly, we posit a compelling argument that
calls for a novel perspective on algorithmic recourse extending our
focus from single individuals to groups. Secondly, we introduce an
experimental framework extending previous work by
\protect\hyperlink{ref-altmeyer2022CounterfactualExplanations}{{[}6{]}},
which enables us to study macrodynamics of algorithmic recourse through
simulations. To this end we propose a number of novel evaluation metrics
that can be used to qunatify and benchmark the macrodynmics introduced
by different counterfactual generators. Thirdly, we use this framework
to provide a first in-depth analysis of endogenous recourse dynamics
induced by various popular counterfactual generators including
\protect\hyperlink{ref-wachter2017counterfactual}{{[}5{]}},
\protect\hyperlink{ref-schut2021generating}{{[}7{]}},
\protect\hyperlink{ref-joshi2019towards}{{[}8{]}},
\protect\hyperlink{ref-mothilal2020explaining}{{[}9{]}} and
\protect\hyperlink{ref-antoran2020getting}{{[}10{]}}. Finally, we
discuss what drives endogenous dynamics and strategies to mitigate them.

The remainder of the paper is structured as follows:
Section~\ref{sec-related} places our work in the broader context of
related literature. Section~\ref{sec-method} presents our methodology
and data. Section~\ref{sec-empirical} presents our empirical findings
which are then discussed in the broader context of the literature in
Section~\ref{sec-discussion}. We also point to some of the limitations
or our work as well as avenues for future research in
Section~\ref{sec-limit}. Finally, Section~\ref{sec-conclusion}
concludes.

@@

Many approaches for the generation of algorithmic recourse have been
described in the literature. An October 2020 survey by Karimi et
al.~layed out 60 algorithms that have been proposed since 2014
\protect\hyperlink{ref-karimi2020survey}{{[}11{]}}. Another survey
published around the same time by Verma et al.~described 29 algorithms
\protect\hyperlink{ref-verma2020counterfactual}{{[}12{]}}. Different
approaches vary primarily in terms of the objective functions they
impose, how they optimize said objective (from brute force through
gradient-based approaches to graph traversal algorithms), and how the
ensure that certain requirements for CE are met. Regarding the latter,
the literature has produced an extensive list of desiderata each
addressing different needs. To name but a few, we are interested in
generating counterfactuals that close
\protect\hyperlink{ref-wachter2017counterfactual}{{[}5{]}}, actionable
\protect\hyperlink{ref-ustun2019actionable}{{[}13{]}}, realistic
\protect\hyperlink{ref-schut2021generating}{{[}7{]}}, sparse, diverse
\protect\hyperlink{ref-mothilal2020explaining}{{[}9{]}} and if possible
causally founded
\protect\hyperlink{ref-karimi2021algorithmic}{{[}14{]}}.

Efforts so far have largely been directed at improving the quality of
counterfactual explanations within a static context: given some
pre-trained classifier \(M: \mathcal{X} \mapsto \mathcal{Y}\) we are
interested in generating one or multiple meaningful counterfactual
explanations for some individual characterized by \(x_i\). The ability
of counterfactual explanations to handle dynamics like data and model
shifts remains a largely unexplored research challenge at this point
\protect\hyperlink{ref-verma2020counterfactual}{{[}12{]}}. We have been
able to identify only one recent work that considers the implications of
\textbf{exogenous} domain and model shifts in the context of AR
\protect\hyperlink{ref-upadhyay2021towards}{{[}15{]}}. Exogenous shifts
are strictly of external origin. For example, they might stem from data
correction, temporal shifts or geospatial changes
\protect\hyperlink{ref-upadhyay2021towards}{{[}15{]}}. The authors of
\protect\hyperlink{ref-upadhyay2021towards}{{[}15{]}} propose framework
for algorithmic recourse (ROAR) that evidently improves robustness to
such exogenous shifts.

@@

Much attention has been paid to the detection of dataset shifts --
situations where the distribution of data changes over time. Rabanser et
al.~suggest a framework to detect data drift from a minimal number of
samples through the application of two-sample tests
\protect\hyperlink{ref-rabanser2019failing}{{[}16{]}}. This task is a
generalization of the anomaly detection problem for large datasets,
which aims to answer the question if two sets of samples could have been
generated from the same probability distribution. Numerous approaches to
anomaly detection have been summarized
\protect\hyperlink{ref-chandola2009anomaly}{{[}17{]}}. Another
well-established research topic is that of concept drift -- situations
where external variables influence the patterns between the input and
the output of a model
\protect\hyperlink{ref-widmer1996learning}{{[}18{]}}. For instance, Gama
et al.~offer a review of the adaptive learning techniques which can
handle concept drift \protect\hyperlink{ref-gama2014survey}{{[}19{]}}.
Less previous work is available on the related topic of model drift -
changes in model performance over time. Nelson et al.~review how
resistant different machine learning models are to the model drift
\protect\hyperlink{ref-nelson2015evaluating}{{[}20{]}}. Ackerman et
al.~offer a method to detect changes in model performance when ground
truth is not available
\protect\hyperlink{ref-ackerman2021machine}{{[}21{]}}.

In the context of algorithmic recourse, domain and model shifts were
first brought up by the authors behind ROAR
\protect\hyperlink{ref-upadhyay2021towards}{{[}15{]}}. In their work
they refer to model shifts as simply any perturbation \(\Delta\) to the
parameters of the model in question: \(M\). While this also sets the
baseline for our analysis here, it is worth noting that in
\protect\hyperlink{ref-upadhyay2021towards}{{[}15{]}} these
perturbations are mechanically introduced. In contrast we are interested
in quantifying model shifts that arise endogenously as part of a dynamic
recourse process. In addition to quantifying the magnitude of shifts
\(\Delta\), we aim to also analyse the characteristics of changes to the
model, such as the position of the decision boundary and the overall
decisiveness of the model. We have not been able to identify previous
work on this topic.

@@

Despite the large and growing number of approaches to counterfactual
search there have been surprisingly few benchmark studies that compare
different methodologies. This may be partially due to limited software
availability in this space. Recent work has started to address this gap:
firstly, \protect\hyperlink{ref-de2021framework}{{[}22{]}} run a large
benchmarking study using different algorithmic aproaches and numerous
tabular datasets; secondly,
\protect\hyperlink{ref-pawelczyk2021carla}{{[}4{]}} introduce a Python
framework that can be used to apply and benchmark different methodology;
finally,
\protect\hyperlink{ref-altmeyer2022CounterfactualExplanations}{{[}6{]}}
provides an extensible and language-agnostic implementation in Julia.
For the experiments conducted in this work we relied on the latter.

@@

Here \(\mathbf{s}^\prime=\left\{s_k^\prime\right\}_K\) is the stacked
\(K\)-dimensional array of counterfactual states and
\(f: \mathcal{S} \mapsto \mathcal{X}\) maps from the counterfactual
state space to the feature space. In the case of the baseline
counterfactual generator
\protect\hyperlink{ref-wachter2017counterfactual}{{[}5{]}} \(f\) is just
the idendity function and the number of counterfactuals \(K\) is equal
to one. This generator, which we shall refer to as \textbf{Wachter} in
the following, shall serve as the baseline against which all other
gradient-based methodologies will be compared. In particular, we include
include the following generator in our benchmarking exercises: REVISE
\protect\hyperlink{ref-joshi2019towards}{{[}8{]}}, CLUE
\protect\hyperlink{ref-antoran2020getting}{{[}10{]}}, DICE
\protect\hyperlink{ref-mothilal2020explaining}{{[}9{]}} and a greedy
approach that relies on probabilistic models
\protect\hyperlink{ref-schut2021generating}{{[}7{]}}.

Both \textbf{REVISE} and \textbf{CLUE} search counterfactuals in some
latent embedding \(S \subset \mathcal{S}\) instead of the feature space
directly. The latent embedding is learned by a separate generative model
that is tasked with learning the data generating process (DGP) of \(X\).
In this case \(f\) in Equation~\ref{eq-general} corresponds to the
decoder part of the generative model, in other words the deterministic
function that maps back from the latent embedding to the feature space.
Provided the generative model is well-specified, traversing the latent
embedding typically results in realistic and plausible counterfactuals,
because they are implicitly generated by the (learned) DGP
\protect\hyperlink{ref-joshi2019towards}{{[}8{]}}. CLUE distinguishes
itself from REVISE and other counterfactual generators in that it aims
to minimize the predictive uncertainty of the model in question \(M\).
To quantify predictive uncertainty the authors rely on entropy estimates
for probabilistic models. The \textbf{Greedy} approach proposed by
\protect\hyperlink{ref-schut2021generating}{{[}7{]}} also works with the
subclass of models \(\tilde{\mathcal{M}}\subset\mathcal{M}\) that can
produce predictive uncertainty estimates. The authors show that in this
setting the complexity penalty \(h(\cdot)\) in Equation~\ref{eq-general}
is redundant and meaningful counterfactuals can be generated in a fast
and efficient manner through a modified Jacobian-based Saliency Map
Attack (JSMA). Finally, \textbf{DICE} distinguishes itself from all
other generators considered here in that it aims to generate a diverse
set of \(K>1\) counterfactuals. To this end the authors use a complexity
penalty \(h(\mathbf{s}^\prime)\) that favours diverse outcomes, in the
sense that \(s_1, ... , s_K\) look as different from each other as
possible.

@@

We have chosen to work with both synthetic and real-world datasets.
Using synthetic data allows us to impose distributional properties that
may affect the resulting recourse dynamics. Following
\protect\hyperlink{ref-upadhyay2021towards}{{[}15{]}} we generate
synthetic data in \(\mathbb{R}^2\) to also allow for a visual
interpretation of the results. Real-world data is used in order to
assess if endogenous dynamics also occur in higher-dimensional settings.

@@

Ex-ante we expect to see that Wachter will create a new cluster of
counterfactual instances in the proximity of the initial decision
boundary. Thus, the choice of a black-box model may have an impact on
the paths of the recourse. For generators that use latent space search
(\protect\hyperlink{ref-joshi2019towards}{{[}8{]}},
\protect\hyperlink{ref-antoran2020getting}{{[}10{]}}) or rely on (and
have access to) probabilistic models
(\protect\hyperlink{ref-antoran2020getting}{{[}10{]}},
\protect\hyperlink{ref-schut2021generating}{{[}7{]}}) we expect that
counterfactuals will end up in regions of the target domain that are
densely populated by training samples. Finally, we expect to see the
counterfactuals generated by DiCE to be uniformly spread around the
feature space inside the target class.

@@

As our baseline for quantifying model shifts we measure perturbations to
the model parameters at each point in time \(t\) following
\protect\hyperlink{ref-upadhyay2021towards}{{[}15{]}}. We define
\(\Delta=||\theta_{t+1}-\theta_{t}||^2\), that is the euclidean distance
between the vectors of parameters before and after retraining the model
\(M\). We shall refer to this baseline metric simply as
\textbf{Perturbations}.

@@

The experimental setup proposed here is designed to mimic a real-world
recourse process in a simple fashion. In practice, models are in fact
updated on a regular basis
\protect\hyperlink{ref-upadhyay2021towards}{{[}15{]}}. We also find it
plausible to assume that the implementation of recourse happens
periodically for different individuals, rather that all at once at time
\(t=0\). That being said, our experimental design is a vast
over-simplification of potential real-world scenarios. In practice, any
endogenous shifts that may occur can be expected to be entangled with
exogenous shifts of the nature investigated in
\protect\hyperlink{ref-upadhyay2021towards}{{[}15{]}}. We also make
implicit assumptions about the utility functions of the involved agents
that may well be too simple: individuals seeking recourse are assumed to
always implement the proposed counterfactual explanations; conversely,
the agent in charge of the model \(M\) is assumed to always treat
individuals that have implemented valid recourse as if they were truly
now in the target class. Relating this back to the consumer credit
example, we assume that the would-be borrowers are always willing and
able to implement recourse and the bank is always willing to provide
credit as would-be borrowers move across the decision boundary. In
practice it is doubtful that agents behave according to such simple
rules. Nonetheless, we think that our simple framework offers a starting
point for future work on recourse dynamics (both endogenous and
exogenous dynamics).

@@

\leavevmode\vadjust pre{\hypertarget{ref-altmeyer2022CounterfactualExplanations}{}}%
\CSLLeftMargin{{[}6{]} }%
\CSLRightInline{P. Altmeyer, \emph{{CounterfactualExplanations.jl - a
Julia package for Counterfactual Explanations and Algorithmic
Recourse}}. 2022. Available:
\url{https://github.com/pat-alt/CounterfactualExplanations.jl}}

\leavevmode\vadjust pre{\hypertarget{ref-schut2021generating}{}}%
\CSLLeftMargin{{[}7{]} }%
\CSLRightInline{L. Schut \emph{et al.}, {``Generating interpretable
counterfactual explanations by implicit minimisation of epistemic and
aleatoric uncertainties,''} in \emph{International conference on
artificial intelligence and statistics}, 2021, pp. 1756--1764.}

\leavevmode\vadjust pre{\hypertarget{ref-joshi2019towards}{}}%
\CSLLeftMargin{{[}8{]} }%
\CSLRightInline{S. Joshi, O. Koyejo, W. Vijitbenjaronk, B. Kim, and J.
Ghosh, {``Towards realistic individual recourse and actionable
explanations in black-box decision making systems,''} \emph{arXiv
preprint arXiv:1907.09615}, 2019.}

\leavevmode\vadjust pre{\hypertarget{ref-mothilal2020explaining}{}}%
\CSLLeftMargin{{[}9{]} }%
\CSLRightInline{R. K. Mothilal, A. Sharma, and C. Tan, {``Explaining
machine learning classifiers through diverse counterfactual
explanations,''} in \emph{Proceedings of the 2020 conference on
fairness, accountability, and transparency}, 2020, pp. 607--617.}

\leavevmode\vadjust pre{\hypertarget{ref-antoran2020getting}{}}%
\CSLLeftMargin{{[}10{]} }%
\CSLRightInline{J. Antorán, U. Bhatt, T. Adel, A. Weller, and J. M.
Hernández-Lobato, {``Getting a clue: A method for explaining uncertainty
estimates,''} \emph{arXiv preprint arXiv:2006.06848}, 2020.}

\leavevmode\vadjust pre{\hypertarget{ref-karimi2020survey}{}}%
\CSLLeftMargin{{[}11{]} }%
\CSLRightInline{A.-H. Karimi, G. Barthe, B. Schölkopf, and I. Valera,
{``A survey of algorithmic recourse: Definitions, formulations,
solutions, and prospects,''} \emph{arXiv preprint arXiv:2010.04050},
2020.}

\leavevmode\vadjust pre{\hypertarget{ref-verma2020counterfactual}{}}%
\CSLLeftMargin{{[}12{]} }%
\CSLRightInline{S. Verma, J. Dickerson, and K. Hines, {``Counterfactual
explanations for machine learning: A review,''} \emph{arXiv preprint
arXiv:2010.10596}, 2020.}

\leavevmode\vadjust pre{\hypertarget{ref-ustun2019actionable}{}}%
\CSLLeftMargin{{[}13{]} }%
\CSLRightInline{B. Ustun, A. Spangher, and Y. Liu, {``Actionable
recourse in linear classification,''} in \emph{Proceedings of the
conference on fairness, accountability, and transparency}, 2019, pp.
10--19.}

\leavevmode\vadjust pre{\hypertarget{ref-karimi2021algorithmic}{}}%
\CSLLeftMargin{{[}14{]} }%
\CSLRightInline{A.-H. Karimi, B. Schölkopf, and I. Valera,
{``Algorithmic recourse: From counterfactual explanations to
interventions,''} in \emph{Proceedings of the 2021 ACM conference on
fairness, accountability, and transparency}, 2021, pp. 353--362.}

\leavevmode\vadjust pre{\hypertarget{ref-upadhyay2021towards}{}}%
\CSLLeftMargin{{[}15{]} }%
\CSLRightInline{S. Upadhyay, S. Joshi, and H. Lakkaraju, {``Towards
robust and reliable algorithmic recourse,''} \emph{arXiv preprint
arXiv:2102.13620}, 2021.}
